### 2013-10-01 src2/arbb/paperThesis.tex
apa

This is a compiler backend research challenge.
The {\em skeletons} method mentioned in
Section~\ref{sec:accelerate-arbb} is one approach to this problem, as
are the optimizations studied in the Copperhead \cite{copperhead} and
Obsidian \cite{OBSIDIAN-TECHREPORT} projects.
%
On the other hand, systems that rely on advanced optimizations
typically suffer to some extent from performance-predictability
problems.  Thus achieving portable, predictable performance on a wide range
of architectures---even for the simplest target languages---will
be the subject of much future work.

>>>>

This is a compiler backend research challenge.
The {\em skeletons} method mentioned in
Section~\ref{sec:accelerate-arbb} is one approach to this problem, as
are the optimizations studied in the Copperhead \cite{copperhead} and
Obsidian \cite{JSLIC} projects.
%
On the other hand, systems that rely on advanced optimizations
typically suffer to some extent from performance-predictability
problems.  Thus achieving portable, predictable performance on a wide range
of architectures---even for the simplest target languages---will
be the subject of much future work.

### 2013-10-01 src2/csort/Obsidian.tex
apa

\section{Obsidian} 
\label{sec:Obsidian} 
%\FloatBarrier

@@

\subsection{A brief history of Obsidian} 
\label{sec:OBSHist}

@@

The Obsidian implementation of pull arrays was influenced by Pan's representation 
of images as functions from coordinates to colour values \cite{ELLIJFP}. A pull array 
is a function from index to value and an associated length.

Push arrays were added to Obsidian first in reference \cite{PUSH}. A push array
specifies where elements are to end up in a result array, rather than where 
they come from (as in a pull array). In essence, a push array is a
computation which writes an array to memory, but is parameterised on
the {\em write function} which writes a single element into memory.
The push array function can then 
use the write function zero, one, or several times.

@@

%The GPUs targeted by Obsidian has an hierarchical structure 
%in more senses than just its memory system. Threads are only 
%allowed to communicate using local shared memory in small groups 
%(thread blocks). 
%To match the hierarchal architecture of GPUs, Obsidian has local arrays 
%of type {\tt Pull a} and {\tt Push a}. These local arrays are used in the 
%implementation of programs that can be run by a block, in Obsidian these 
%programs are represented by a data type called {\tt BProgram}. There 
%are also global arrays types {\tt GlobPull a} and {\tt GlobPush a} that 
%are used in the implementation of programs executed by more than one MP, 
%these programs are represented by a {\tt GProgram} data type.   
%and global arrays of 
%types {\tt GlobPull a} and {\tt GlobPush a}.
%Global arrays are used when simple permutations of a large input array is 
%needed. Here it is important to that no communication needing 
%synchronisations occurs. Global arrays are also used as the inputs 
%and outputs of a top level Obsidian program (a program you want to generate runnable CUDA code from). In the cases where threads need to cooperate 
%on solving a task the local arrays should be used. 
\subsection{Obsidian programming example} 
\label{sec:sklansky}
\FloatBarrier

@@

%We want to be able to sort large arrays and to do this we need an 
%implementation of prefix sum that is larger (on more elements) than what 
%fits in a block. 
We need to operate on arrays that are too long to fit in a single block.
%This large prefix sum is implemented as shown in
The resulting prefix sum is implemented as shown in reference 
\cite{LargeScan}, see figure \ref{fig:prefixsum}. There are three steps 
involved. First local prefix sums are computed; the rightmost elements elements are also stored in a separate, auxiliary, array. Following this, 
the prefix sum of the auxiliary array is  computed. The prefix sums used 
in phase A and B of the 
figure can both be generated from the same Obsidian program. Finally the 
local prefix sums and the result of the auxiliary prefix sum are combined, 
forming a large prefix sum over the entire array.

@@

\subsection{A note about generated code}

>>>>

\subsection{Obsidian} 
\label{sec:Obsidian} 
%\FloatBarrier

@@

\subsubsection{A brief history of Obsidian} 
\label{sec:OBSHist}

@@

The Obsidian implementation of pull arrays was influenced by Pan's representation 
of images as functions from coordinates to colour values \cite{PAN}. A pull array 
is a function from index to value and an associated length.

Push arrays were added to Obsidian first in reference \cite{Obsidian-Expressive}. A push array
specifies where elements are to end up in a result array, rather than where 
they come from (as in a pull array). In essence, a push array is a
computation which writes an array to memory, but is parameterised on
the {\em write function} which writes a single element into memory.
The push array function can then 
use the write function zero, one, or several times.

@@

%The GPUs targeted by Obsidian has an hierarchical structure 
%in more senses than just its memory system. Threads are only 
%allowed to communicate using local shared memory in small groups 
%(thread blocks). 
%To match the hierarchal architecture of GPUs, Obsidian has local arrays 
%of type {\tt Pull a} and {\tt Push a}. These local arrays are used in the 
%implementation of programs that can be run by a block, in Obsidian these 
%programs are represented by a data type called {\tt BProgram}. There 
%are also global arrays types {\tt GlobPull a} and {\tt GlobPush a} that 
%are used in the implementation of programs executed by more than one MP, 
%these programs are represented by a {\tt GProgram} data type.   
%and global arrays of 
%types {\tt GlobPull a} and {\tt GlobPush a}.
%Global arrays are used when simple permutations of a large input array is 
%needed. Here it is important to that no communication needing 
%synchronisations occurs. Global arrays are also used as the inputs 
%and outputs of a top level Obsidian program (a program you want to generate runnable CUDA code from). In the cases where threads need to cooperate 
%on solving a task the local arrays should be used. 
\subsubsection{Obsidian programming example} 
\label{sec:sklansky}
\FloatBarrier

@@

%We want to be able to sort large arrays and to do this we need an 
%implementation of prefix sum that is larger (on more elements) than what 
%fits in a block. 
We need to operate on arrays that are too long to fit in a single block.
%This large prefix sum is implemented as shown in
The resulting prefix sum is implemented as shown in reference 
\cite{ScanCUDA}, see figure \ref{fig:prefixsum}. There are three steps 
involved. First local prefix sums are computed; the rightmost elements elements are also stored in a separate, auxiliary, array. Following this, 
the prefix sum of the auxiliary array is  computed. The prefix sums used 
in phase A and B of the 
figure can both be generated from the same Obsidian program. Finally the 
local prefix sums and the result of the auxiliary prefix sum are combined, 
forming a large prefix sum over the entire array.

@@

\subsubsection{A note about generated code}

### 2013-10-01 src2/csort/intro.tex
apa

These algorithms are explored in Obsidian \cite{JSLIC,PUSH}, a domain
specific language targeting GPU programming. The goal of Obsidian is
to strike a balance between high-level constructs and low-level control. 
We want to provide the programmer with a set of tools for low-level 
experimentation with the details that influence performance when 
implementing GPU kernels. The counting sort
case study shows that Obsidian generates competitive kernels with
relatively little programmer effort. That Obsidian is an embedded language 
allows us to rapidly experiment with the addition of features and with varying 
programming idioms. The version used in this case study
adds global arrays and atomic instructions to Obsidian, see section 
\ref{sec:OBSHist}.

@@

Sorting has applications in the computer graphics field \cite{sintorn}. 
Example instances of sorting and duplicate element removal in computer graphics 
can be identified in references \cite{Olsson} and \cite{Karras}. Another example 
of where sorting and the removal of duplicates have applications is 
in databases \cite{REMOVEDUPS}.

@@

Obsidian is a high-level embedded language for GPU
programming. Other such approaches include Accelerate
\cite{Acc} and Nikola \cite{NIKOLA}. Accelerate and Nikola both take an even
higher level approach compared to Obsidian and abstract more from GPU
details. Another example of providing a high-level interface to
programming the GPU is the C++ library Thrust \cite{THRUST}.
%\cite{FELDSPAR}

>>>>

These algorithms are explored in Obsidian \cite{JSLIC,Obsidian-Expressive}, a domain
specific language targeting GPU programming. The goal of Obsidian is
to strike a balance between high-level constructs and low-level control. 
We want to provide the programmer with a set of tools for low-level 
experimentation with the details that influence performance when 
implementing GPU kernels. The counting sort
case study shows that Obsidian generates competitive kernels with
relatively little programmer effort. That Obsidian is an embedded language 
allows us to rapidly experiment with the addition of features and with varying 
programming idioms. The version used in this case study
adds global arrays and atomic instructions to Obsidian, see section 
\ref{sec:OBSHist}.

@@

Sorting has applications in the computer graphics field \cite{SINTORNSORT}. 
Example instances of sorting and duplicate element removal in computer graphics 
can be identified in references \cite{Olsson} and \cite{Karras}. Another example 
of where sorting and the removal of duplicates have applications is 
in databases \cite{REMOVEDUPS}.

@@

Obsidian is a high-level embedded language for GPU
programming. Other such approaches include Accelerate
\cite{ACCELERATEDAMP11} and Nikola \cite{NIKOLA}. Accelerate and Nikola both take an even
higher level approach compared to Obsidian and abstract more from GPU
details. Another example of providing a high-level interface to
programming the GPU is the C++ library Thrust \cite{THRUST}.
%\cite{FELDSPAR}

### 2013-10-01 src2/embarbb/implementation.tex
apa

The expression data type used in EmbArBB is untyped (not using a GADT). 
A typed interface to the language is supplied using the same phantom types 
method as used in {\em pan}~\cite{ELLIJFP} and many other Haskell embedded 
languages since then. The choice to use an untyped {\tt Expr} datatype was 
based on the wish to keep the backend (ArBB code generation) as simple as 
possible.

>>>>

The expression data type used in EmbArBB is untyped (not using a GADT). 
A typed interface to the language is supplied using the same phantom types 
method as used in {\em pan}~\cite{COMPILEEDSL} and many other Haskell embedded 
languages since then. The choice to use an untyped {\tt Expr} datatype was 
based on the wish to keep the backend (ArBB code generation) as simple as 
possible.

### 2013-10-01 src2/ifl/related.tex
apa

\subsection{Discussion} \label{sec:disc}
\subsubsection{Our influences}
As mentioned above, our earlier work on Lava has provided the inspiration for
the combinator-oriented or hardware-like style of programming that we are exploring in Obsidian.
On the other hand, the {\em implementation} of Obsidian has been much influenced by
{Pan}, an embedded language for image synthesis developed by 
Conal Elliot~\cite{PAN}. Because of the computational complexity of image generation, C 
code is generated. This C code can then be compiled by an optimising compiler. 
Many ideas from the paper 
``Compiling Embedded Languages'', describing the implementation of Pan have been 
used in the implementation of Obsidian~\cite{Elliott03:CompileDSEL-JFP}.

\subsubsection{Related work on GPU and GPGPU programming languages}
We cannot attempt an exhaustive description of GPU programming languages here, but refer the reader to a recent PhD thesis by Philipp Lucas, which contains an enlightening survey~\cite{Lucas08}. Lucas distinguishes carefully between languages (such as CG and HLSL) that aim to raise the level of abstraction at which graphics-oriented GPU programs are written, and those that attempt to abstract the entire GPU, and so must also provide a means to express the placing of programs on the GPU, feeding such programs with data, reading the results back to the CPU, and so on, as well as deciding to what extent the programmer should be involved
in stipulating those tasks.
In the first group of graphics-oriented languages, we include PyGPU and Vertigo.
PyGPU is a language for image processing embedded in Python~\cite{LUNDPAPER}. 
PyGPU uses the introspective abilities of Python and thus bypasses
the need to implement new loop structures and conditionals for the embedded 
language. In Python it is possible to access the bytecode of a function and 
from that extract information about loops and conditionals. 
Programs written in PyGPU can be compiled and run on a GPU. 
Vertigo is another embedded language by Conal Elliot~\cite{VERTIGO}. It
is a language for 3D graphics that targets the DirectX 8.1 shader model, and
can be used to describe geometry, shaders and to generate textures. 
%Each sublanguage is given formal semantics \cite{VERTIGO}. From programs 
%written in Vertigo, assembly language programs are generated for execution on a 
%GPU.

>>>>

\subsection{Discussion} \label{sec:disc}
\subsubsection{Our influences}
As mentioned above, our earlier work on Lava has provided the inspiration for
the combinator-oriented or hardware-like style of programming that we are exploring in Obsidian.
On the other hand, the {\em implementation} of Obsidian has been much influenced by
{Pan}, an embedded language for image synthesis developed by 
Conal Elliot~\cite{PAN}. Because of the computational complexity of image generation, C 
code is generated. This C code can then be compiled by an optimising compiler. 
Many ideas from the paper 
``Compiling Embedded Languages'', describing the implementation of Pan have been 
used in the implementation of Obsidian~\cite{COMPILEEDSL}.

\subsubsection{Related work on GPU and GPGPU programming languages}
We cannot attempt an exhaustive description of GPU programming languages here, but refer the reader to a recent PhD thesis by Philipp Lucas, which contains an enlightening survey~\cite{Lucas08}. Lucas distinguishes carefully between languages (such as CG and HLSL) that aim to raise the level of abstraction at which graphics-oriented GPU programs are written, and those that attempt to abstract the entire GPU, and so must also provide a means to express the placing of programs on the GPU, feeding such programs with data, reading the results back to the CPU, and so on, as well as deciding to what extent the programmer should be involved
in stipulating those tasks.
In the first group of graphics-oriented languages, we include PyGPU and Vertigo.
PyGPU is a language for image processing embedded in Python~\cite{PyGPU}. 
PyGPU uses the introspective abilities of Python and thus bypasses
the need to implement new loop structures and conditionals for the embedded 
language. In Python it is possible to access the bytecode of a function and 
from that extract information about loops and conditionals. 
Programs written in PyGPU can be compiled and run on a GPU. 
Vertigo is another embedded language by Conal Elliot~\cite{VERTIGO}. It
is a language for 3D graphics that targets the DirectX 8.1 shader model, and
can be used to describe geometry, shaders and to generate textures. 
%Each sublanguage is given formal semantics \cite{VERTIGO}. From programs 
%written in Vertigo, assembly language programs are generated for execution on a 
%GPU.

### 2013-10-01 src2/papp/paperThesis.tex
apa

Obsidian is a domain specific language for general purpose programming on 
GPUs (GPGPU), capable of generating code for modern NVIDIA GPUs. 
In~\cite{JSMSKC_IFL08} we describe a version of Obsidian that uses a monadic 
interface. In the current version of Obsidian the monad has been replaced 
by another data structure, closely related to Arrows~\cite{Arrows}, representing GPU programs. Although the details are 
left out because of space limitations, the use of this GPU program representation 
can be seen in subsection~\ref{sec:GPUPrograms}. For a description of the implementation 
see~\cite{JSTECH} and for more general information on embedded language 
implementation see~\cite{Elliott03:CompileDSEL-JFP}.

@@

Sorting is a popular function to place on the GPU, see~\cite{SORT1,SORT3}.
{\em Odd-Even merge sort}, due to Batcher, uses a merger component called the Odd-Even merger. This 
merger merges two sorted arrays. In Obsidian this merger and sorter 
can be implemented using the combinators {\tt two}, {\tt ilv}, {\tt evens}
and {\tt odds}. For a detailed description of these combinators see
~\cite{LAVASORTER}.

@@

The Sklansky kernels given above can be used in an algorithm that computes 
the parallel prefix of a large array. This is done in the same way as in
~\cite{harris07parallel}.
The table below shows the results of using the different parallel prefix 
kernels from above in an algorithm for computing the prefix sums of 
$2^{20}$ elements.

@@

The table above shows the running times of six different Sklansky kernels. 
All runtime measurements where performed on an NVIDIA 9800GX2 using one GPU. 
The code labeled {\em Hand Optimised} was written directly in CUDA. 
This kernel was the result of two afternoons of optimisation effort by 
two people. The three following kernels {\em sklansky} to {\em sklansky2} are 
generated from the given Obsidian programs. The version called {\em sklansky3} 
is identical to {\tt sklansky2} except is in-place. Even the very first version, 
{\em sklansky} performs reasonably well, but by using the experimental features 
the performance can be pushed quite close to the hand optimised version.  
The fastest kernel, called {\em NVIDIA SDK}, is the one supplied 
with the CUDA SDK. This kernel is based on the Brent-Kung network and its 
implementation is shown in~\cite{harris07parallel}. The {\tt NVIDIA SDK} 
kernel is highly optimized with regards to its memory access pattern and 
so the code is considerably more complicated than that of any of the Obsidian parallel-prefix kernels.

>>>>

Obsidian is a domain specific language for general purpose programming on 
GPUs (GPGPU), capable of generating code for modern NVIDIA GPUs. 
In~\cite{JSMSKC_IFL08} we describe a version of Obsidian that uses a monadic 
interface. In the current version of Obsidian the monad has been replaced 
by another data structure, closely related to Arrows~\cite{Arrows}, representing GPU programs. Although the details are 
left out because of space limitations, the use of this GPU program representation 
can be seen in subsection~\ref{sec:GPUPrograms}. For a description of the implementation 
see~\cite{JSTECH} and for more general information on embedded language 
implementation see~\cite{COMPILEEDSL}.

@@

Sorting is a popular function to place on the GPU, see~\cite{SINTORNSORT,SORT3}.
{\em Odd-Even merge sort}, due to Batcher, uses a merger component called the Odd-Even merger. This 
merger merges two sorted arrays. In Obsidian this merger and sorter 
can be implemented using the combinators {\tt two}, {\tt ilv}, {\tt evens}
and {\tt odds}. For a detailed description of these combinators see
~\cite{LAVASORTER}.

@@

The Sklansky kernels given above can be used in an algorithm that computes 
the parallel prefix of a large array. This is done in the same way as in
~\cite{ScanCUDA}.
The table below shows the results of using the different parallel prefix 
kernels from above in an algorithm for computing the prefix sums of 
$2^{20}$ elements.

@@

The table above shows the running times of six different Sklansky kernels. 
All runtime measurements where performed on an NVIDIA 9800GX2 using one GPU. 
The code labeled {\em Hand Optimised} was written directly in CUDA. 
This kernel was the result of two afternoons of optimisation effort by 
two people. The three following kernels {\em sklansky} to {\em sklansky2} are 
generated from the given Obsidian programs. The version called {\em sklansky3} 
is identical to {\tt sklansky2} except is in-place. Even the very first version, 
{\em sklansky} performs reasonably well, but by using the experimental features 
the performance can be pushed quite close to the hand optimised version.  
The fastest kernel, called {\em NVIDIA SDK}, is the one supplied 
with the CUDA SDK. This kernel is based on the Brent-Kung network and its 
implementation is shown in~\cite{ScanCUDA}. The {\tt NVIDIA SDK} 
kernel is highly optimized with regards to its memory access pattern and 
so the code is considerably more complicated than that of any of the Obsidian parallel-prefix kernels.
